Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?

### 2017-09-28 paper.tex
Update on Overleaf.

In order to test for the influence of spatial initial conditions on model outputs, we use a systematic method to compare phase diagrams. We define a phase diagram as a vector of  outputs\comment[FL]{Juste, peut-être des choses qui sont mal dites? : par exemple, dans diagramme de phase est-ce qu'on entend pas une "réification" de catégories - des clusters par ex? Et du coup est-ce que cela ne manque pas là?}.

@@

For Sugarscape, we explore a grid of a basic parameter space of the model, which is composed of three dimensions: the population of agents $P\in \left[10;510\right]$, the minimal initial agent resource $s_{-}\in \left[10;100\right]$ and the maximal initial agent resource $s_{+}\in \left[110;200\right]$. Each parameter is binned into 10 values, giving 1000 parameter points. We run 50 repetitions for each configuration, what yield reasonable convergence properties. The initial spatial configuration varies across 50 different grids, generated by sampling meta-parameters for the generator in a LHS. We did not use the clustered grids to test the flexibility of our framework, which is demonstrated in this case by a direct sequential coupling of the generator and the model. The full experiment thus equates to 2,500,000 simulations (1000 parameter combinations x 50 density grids x 50 replications).

@@

\begin{table*}[htbp]
\caption{\textbf{Regression of the segregation level of Schelling simulation with order parameters and type of city grid.} *** means that the estimate is significant.\label{tab:regressionSchelling}}
\begin{adjustwidth}{-0.5cm}{-0.5cm}
\centering
\begin{tabular}{|m{2cm}|ll|ll|ll|}
\hline
Simulation outcome by segregation index: & \multicolumn{2}{c|}{\textbf{Dissimilarity}} & \multicolumn{2}{c|}{\textbf{Entropy}} & \multicolumn{2}{c|}{\textbf{Moran's I}} \\ \hline
\textbf{Intercept}                       & -0.007 ***  & -0.03 ***                     & -0.028 ***        & -0.041 ***        & 0.127 ***        & 0.129 ***            \\ \hline
\textbf{Similarity Wanted (S)}           & -0.007 ***  & -0.007 ***                    & -0.007 ***        & -0.007 ***        & -0.005 ***       & -0.005 ***           \\ 
\textbf{Vacancy Rate (V)}                & 0.294 ***   & 0.294 ***                     & 0.204 ***         & 0.294 ***         & 0.204 ***        & \textit{0.099}       \\ \hline
\textbf{Density Grid = Polycentric}      &             & 0.038 ***                     &                   & 0.023 ***         &                  & -0.006 ***           \\ 
\textbf{Density Grid = Discontinuous}    &             & 0.03 ***                     &                   & 0.017 ***         &                  & -0.001 ***           \\ \hline
\textbf{R2 (\%)}                         & 15.01       & 15.57  & 7.62              & 7.83              & 2.04             & 2.05                 \\ 
\textbf{\# of observations (sim. runs)}  & 4,500,000   & 4,500,000                     & 4,500,000         & 4,500,000         & 4,500,000        & 4,500,000            \\ \hline
\end{tabular}
\end{adjustwidth}
\end{table*}

\begin{table*}[]
\centering
\caption{\comment[cc]{Resultats Temporaires sur 1/3 des simu}}
\begin{adjustwidth}{-0.5cm}{-0.5cm}
\begin{tabular}{|m{2.5cm}|ll|ll|ll|}
\hline
Simulation outcome by segregation index:    & \multicolumn{2}{c|}{\textbf{Dissimilarity}}   & \multicolumn{2}{c|}{\textbf{Entropy}} & \multicolumn{2}{c|}{\textbf{Moran's I}} \\ \hline
\textbf{Intercept}                          & -0.222 *** & -0.206 ***                       & -0.272 ***        & -0.288 ***        & -0.040 ***           & -0.086 ***               \\ \hline
\textbf{Similarity Wanted (S)}              & 1.426 ***  & 1.426 ***                        & 1.482 ***         & 1.482 ***         & 0.673 ***            & 0.673 ***                \\ 
\textbf{quadratic term ($S^2$)}               & -1.158 *** & -1.158 ***                       & -1.179 ***        & -1.179 ***        & -0.545 ***           & -0.545 ***               \\ 
\textbf{Vacancy Rate (V)}                   & 0.614 ***  & 0.614 ***                        & 0.486 ***         & 0.486 ***         & -0.037 ***           & -0.037 ***               \\ 
\textbf{Minority Index (\%Maj - \%Min)}     & 0.280 ***  & 0.280 ***                        & 0.094 ***         & 0.094 ***         & -0.090 ***           & -0.090 ***               \\ \hline
\textbf{Density Grid = Polycentric}         &            & 0.098 ***                        &                   & 0.070 ***         &                      & 0.002 ***                \\ 
\textbf{Density Grid = Discontinuous}       &            & 0.117 ***                        &                   & 0.082 ***         &                      & -0.001 ***               \\
\textbf{Attraction meta-parameter $\alpha$} &            & -0.055 ***                       &                   & -0.024 ***        &                      & 0.026 ***                \\ 
\textbf{Diffusion meta-parameter $\beta$}   &            & 0.270 ***                        &                   & 0.175 ***         &                      & \textit{-0.00}           \\ \hline
\textbf{R2 (\%)}                            & 34.6       & 38.5                             & 29.1              & 30.8              & 29.8                 & 30.0                     \\ 
\textbf{\# of observations (sim. runs)}     & 1,403,899  & {\color[HTML]{000000} 1,403,899} & 1,403,899         & 1,403,899         & 1,403,899            & 1,403,899                \\ 
\textbf{AIC}                                & -5939.02   & -91944.86                        & 246882.4          & 213145.1          & -2829106             & -2832773                 \\ \hline
\end{tabular}
\end{adjustwidth}
N.B. Moran's I applies to the minority population
\end{table*}

\begin{table*}[]
\centering
\caption{\comment[CC]{Nouveaux Resultats}}
\begin{adjustwidth}{-0.5cm}{-0.5cm}
\begin{tabular}{|m{2.5cm}|ll|ll|ll|}
\hline
Simulation outcome by segregation index:    & \multicolumn{2}{c|}{\textbf{Dissimilarity}}   & \multicolumn{2}{c|}{\textbf{Entropy}} & \multicolumn{2}{c|}{\textbf{Moran's I}} \\ \hline
\textbf{Intercept}                          & -0.212 *** & -0.141 ***                       & -0.254 ***        & -0.208 ***        & -0.036 ***           & -0.061 ***               \\ \hline
\textbf{Similarity Wanted (S)}              & 1.212 ***  & 1.212 ***                        & 1.250 ***         & 1.250 ***         & 0.550 ***            & 0.550 ***                \\ 
\textbf{quadratic term ($S^2$)}               & -0.942 *** & -0.942 ***                       & -0.963 ***        & -0.963 ***        & -0.428 ***           & -0.438 ***               \\ 
\textbf{Vacancy Rate (V)}                   & 0.602 ***  & 0.602 ***                        & 0.453 ***         & 0.453 ***         & -0.027 ***           & -0.027 ***               \\ 
\textbf{Minority Index (\%Maj - \%Min)}     & 0.307å ***  & 0.307 ***                        & 0.130 ***         & 0.130 ***         & -0.067 ***           & -0.067 ***               \\ \hline
\textbf{Density Grid = Polycentric}         &            & 0.087 ***                        &                   & 0.052 ***         &                      & 0.001 ***                \\ 
\textbf{Density Grid = Discontinuous}       &            & 0.111 ***                        &                   & 0.068 ***         &                      & \textit{0.00}              \\
\textbf{Attraction meta-parameter $\alpha$} &            & -0.083 ***                       &                   & -0.053 ***        &                      & 0.014 ***                \\ 
\textbf{Diffusion meta-parameter $\beta$}   &            & 0.323 ***                        &                   & 0.218 ***         &                      & 0.017 ***           \\ \hline
\textbf{R2 (\%)}                            & 30.6       & 34.7                             & 24.1              & 25.6              & 23.9                 & 24.0                    \\ 
\textbf{\# of observations (sim. runs)}     & 2,106,000  & 2,106,000 						 & 2,106,000          & 2,106,000          & 2,106,000             & 2,106,000                \\ 
\textbf{AIC}                                & -70717.68   & -198748.2  						& 208213.8          & 166048.8          & -4385990             & -4387816                 \\ \hline
\end{tabular}
\end{adjustwidth}
N.B. Moran's I applies to the minority population
\end{table*}

>>>>

In order to test for the influence of spatial initial conditions on model outputs, we use a systematic method to compare phase diagrams. We define a phase diagram as the projection of a vector of model outputs in the parameter space. Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values as the median of the final values of an output indicator obtained for the replications of the model initialized with this combination of parameter.\\

@@

For Sugarscape, we explore a grid of a basic parameter space of the model, which is composed of three dimensions: the population of agents $P\in \left[10;510\right]$, the minimal initial agent resource $s_{-}\in \left[10;100\right]$ and the maximal initial agent resource $s_{+}\in \left[110;200\right]$. Each parameter is binned into 10 values, giving 1000 parameter points. We run 50 repetitions for each configuration, what yield reasonable convergence properties. The initial spatial configuration varies across 50 different grids, generated by sampling meta-parameters for the generator in a LHS. We did not use the clustered grids to test the flexibility of our framework, which is demonstrated in this case by a direct sequential coupling of the generator and the model. Indeed, because the density distribution refers to the distribution of resource rather than to the representation of a city structure, we do not need the typology of urban density in this experiment. The full experiment thus equates to 2,500,000 simulations (1000 parameter combinations x 50 density grids x 50 replications).

@@

\begin{table*}[]
\centering
\caption{\textbf{Regression of the segregation level of Schelling simulation with order parameters and type of city grid.} *** means that the estimate is significant.}
\label{tab:regressionSchelling}
\begin{adjustwidth}{-0.5cm}{-0.5cm}
\begin{tabular}{|m{2.5cm}|ll|ll|ll|}
\hline
Simulation outcome by segregation index:    & \multicolumn{2}{c|}{\textbf{Dissimilarity}}   & \multicolumn{2}{c|}{\textbf{Entropy}} & \multicolumn{2}{c|}{\textbf{Moran's I}} \\ \hline
\textbf{Intercept}                          & -0.212 *** & -0.141 ***                       & -0.254 ***        & -0.208 ***        & -0.036 ***           & -0.061 ***               \\ \hline
\textbf{Similarity Wanted (S)}              & 1.212 ***  & 1.212 ***                        & 1.250 ***         & 1.250 ***         & 0.550 ***            & 0.550 ***                \\ 
\textbf{quadratic term ($S^2$)}               & -0.942 *** & -0.942 ***                       & -0.963 ***        & -0.963 ***        & -0.428 ***           & -0.438 ***               \\ 
\textbf{Vacancy Rate (V)}                   & 0.602 ***  & 0.602 ***                        & 0.453 ***         & 0.453 ***         & -0.027 ***           & -0.027 ***               \\ 
\textbf{Minority Index (\%Maj - \%Min)}     & 0.307å ***  & 0.307 ***                        & 0.130 ***         & 0.130 ***         & -0.067 ***           & -0.067 ***               \\ \hline
\textbf{Density Grid = Polycentric}         &            & 0.087 ***                        &                   & 0.052 ***         &                      & 0.001 ***                \\ 
\textbf{Density Grid = Discontinuous}       &            & 0.111 ***                        &                   & 0.068 ***         &                      & \textit{0.00}              \\
\textbf{Attraction meta-parameter $\alpha$} &            & -0.083 ***                       &                   & -0.053 ***        &                      & 0.014 ***                \\ 
\textbf{Diffusion meta-parameter $\beta$}   &            & 0.323 ***                        &                   & 0.218 ***         &                      & 0.017 ***           \\ \hline
\textbf{R2 (\%)}                            & 30.6       & 34.7                             & 24.1              & 25.6              & 23.9                 & 24.0                    \\ 
\textbf{\# of observations (sim. runs)}     & 2,106,000  & 2,106,000 						 & 2,106,000          & 2,106,000          & 2,106,000             & 2,106,000                \\ 
\textbf{AIC}                                & -70717.68   & -198748.2  						& 208213.8          & 166048.8          & -4385990             & -4387816                 \\ \hline
\end{tabular}
\end{adjustwidth}
N.B. Moran's I applies to the minority population
\end{table*}
